Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2018-06-19 writeup/conStruct.tex

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \cite{CavalliSforza1975, treemix},
dimensionality-reduction approaches, 
such as principal components analysis 
\cite{meirmans2009genodive,menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
(e.g., \cite{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}), among others.
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of differentiation,
as that produced by geographic isolation by distance 
\cite{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
(We note that the problem of identifying population clusters is distinct from, 
though of course related to, 
the problem of detecting barriers to gene flow between populations, 
e.g.,  
\cite{BEDASSLE,EEMS,barton2008,Ringbauer2018}.)
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on cluster membership \cite{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.

@@

\paragraph{Data}
The statistical framework of our approach is conceptually similar to that in \cite{Wasser2004}, \cite{BEDASSLE}, and \cite{spacemix},
although we use a somewhat different summary statistic than in this previous work.
The model works with allele frequencies at $L$ unlinked, bi-allelic single nucleotide polymorphisms (SNPs) genotyped across $N$ samples.
Each ``sample'' may be a single individual,  
a collection of individuals from a location, 
or frequencies estimated from pooled sequencing.
From these we compute the \emph{allelic covariance} 
between samples $i$ and $j$, denoted $\widehat{\Omega}_{i,j}$,
as the expected covariance of distinct individual alleles 
chosen from each of the two samples at a random locus.
More precisely, suppose that we pick a random bi-allelic locus uniformly,
pick a random ``reference'' allelic state from the two alleles seen at that locus,
and, in each sample, draw one random allele,
recording $X_i = 1$ if the allele drawn in sample $i$ matches the random reference,
and $X_i = 0$ otherwise.
Then, 
\begin{equation} \label{eqn:first_allelic_cov}
    \widehat{\Omega}_{i,j} = \cov[X_i, X_j] .
\end{equation}
Because we randomly choose the reference allele, 
each $X_i$ behaves marginally as a fair coin --
in particular, $\P\{X_i = 1\} = 1/2$, so $\widehat{\Omega}_{i,i} = 1/4$ for every $i$ --
all information enters through \emph{correlations}.

@@

Within each of these layers,
allele frequencies have positive covariance at geographically close locations,
but this covariance is allowed to decay as geographic distance increases.
This pattern of spatial decay reflects how migration between neighboring demes  
homogenizes allele frequency changes that arise locally due to drift, 
but less effectively homogenizes geographic distant demes,
resulting in a continuous pattern of isolation by distance within each layer.
There is a fixed amount of covariance between layers, irrespective of spatial location.
Within each layer, allele frequencies are expected to change gradually with distance,
but observed frequencies can change abruptly at many loci 
if the proportions of ancestry individuals derive from each layer 
(the ``admixture'' proportions) 
do so as well.

@@

Cross-validation provides a valuable summary of how much explanatory power
is added by spatial structure within each layer, and each additional layer.
However, we remind users that ``statistical significance does not imply real-world significance','
and so small but statistically significant differences between models
should probably not be relied on too strongly.

@@

To calculate layer contributions, we use the following alternative description of our covariance model:
the genomes of any pair of individuals \emph{agree} with some
background probability at a locus,
but this probability of agreement is increased on any segment of genome 
that both have inherited from the same layer 
(the amount it increases depends on how far apart they are
geographically and on the decay of isolation by distance).
We use this characterization to quantify the relative contributions of each layer,
by computing the average contribution to increased probability of agreement
as described in \secref{layer_contribution}). 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

@@

To test the method, we first generated data using the coalescent simulator \texttt{ms} \cite{Hudson2002}.
In each simulation,
we split a single ancestral population into $K$ subpopulations $\tau_{\text{s}}$ units of coalescent time in the past,
and at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonized a
separate $6 \times 6$ square lattice of demes.
Migration on each lattice was to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past, 
we collapsed those $K$ discrete layers into a single grid of demes,
choosing various amounts of admixture from these different layers
(see Fig \ref{sim_setup}). 
We collapsed the layers together using random, spatially autocorrelated admixture proportions 
(see \secref{sim_details}).
We simulated datasets using $K=1$, 2, and 3 layers;
in each simulation we sampled 10,000 unlinked loci from each of 20 haploid individuals from every deme.
We then ran both spatial and nonspatial \texttt{conStruct} analyses on each simulated dataset with $K$ between 1 and 7,
and compared predictive performance of the models using cross-validation with 10 replicates.
For comparison,
we also analyzed each simulated dataset using ADMIXTURE \cite{ADMIXTURE} with $K$ between 1 and 7, 
and compared models using ADMIXTURE's cross-validation procedure with 50 data fold subsets.

@@

The spatial model's better fit is reflected by increased predictive accuracy:
as shown in Fig \ref{sim_xvals},
across all models and choices of $K$, the spatial model is correctly preferred over the nonspatial model.
As desired, predictive accuracy of the spatial model increases until the true value of $K$,
and then plateaus or declines
(Figs \ref{sim_xvals} and \ref{simK1_xval}-\ref{simK3_xval}),
while predictive accuracy of the nonspatial model increases as subsequent clusters are added.
The same holds true for the ADMIXTURE cross-validation results,
in which models that have the largest number of clusters are preferred over all other models,
as shown in Fig \ref{sim_xvals} (vermilion diamonds),
and, in more detail, in Fig \ref{sim_admix_CVerrors}.

@@

\paragraph{Comparison to ADMIXTURE}
Both nonspatial \texttt{conStruct} and ADMIXTURE displayed 
the successive partitioning of space and the clines of admixture seen in the simulation results.
The details of each were somewhat different 
(Fig \ref{populus_nsp_pies} vs. Fig \ref{populus_admixture}), 
and also differed across the replicate analyses.
These differences between runs and methods may be due to 
noise in the different inference algorithms employed,
multi-modality in the likelihood surfaces, or 
to differences in the specifics of model 
(e.g., the priors used in nonspatial \texttt{conStruct}, 
or the fact that ADMIXTURE is modeling each allele's frequency in each cluster, 
rather than the covariance across all alleles).
However, overall, the behavior of both methods was quite similar: 
each recovered the \tri{}/\bals{} split with the first two clusters modeled, 
then, with higher values of $K$, 
used subsequent clusters to subdivide the \tri{} samples 
into geographically restricted foci of cluster membership.
And, both nonspatial \texttt{conStruct} and ADMIXTURE 
strongly favored the most cluster-rich model
(Figs \ref{populus_xvals} and \ref{poplar_admix_CVerrors}).
In contrast, the spatial \texttt{conStruct} model clearly did not favor 
the model with the highest value of $K$, 
and appears to describe patterns of isolation by distance between
demes within \tri{} quite well.

@@

\paragraph{Results from \texttt{conStruct}}
The results partition the sampled bears into two main groups:
one (red) to the east of the Rocky Mountains, 
which also occurs in Alaska,
the other primarily west of the Rockies (blue)  (Fig \ref{bear_K3}a).
The disjointed range of the blue layer likely reflects the fact
that Canada was not sampled, 
and so the red layer may extend
through the intervening (unsampled) northern Great Plains and Canadian Shield, 
with the blue layer presumably then stretching up into British Columbia.

The spatial models have strong statistical support 
up until around $K=5$ or 6 (Fig \ref{bear_xvals}),
but additional spatial layers beyond $K=2$ contribute little to total covariance (Fig \ref{bear_laycon}).
The locations of admixed individuals 
are consistent with a scenario of postglacial expansion from two refugia, 
one in the American Southwest and one in the American Southeast, 
meeting near the Northwest coast of North America and the Cascade Range.
However, lack of any samples from Canada and Mexico, 
and lack of denser sampling across northern North America, 
make more detailed interpretations untrustworthy.
The spatial covariance functions estimated in layers beyond the first two 
take very large values over small spatial lags, 
but decay sharply after that.
This feature, combined with the overall amounts 
and spatial patterns of ancestry in those layers, 
suggests that these layers are describing processes 
that shape genetic variation at local scales, such as inbreeding, 
which affects covariance between individuals within demes, 
but has limited impact on covariance between demes.

\paragraph{Comparison to ADMIXTURE}
Results from the nonspatial model and from the ADMIXTURE analyses 
clearly exhibit the tendency of nonspatial clustering algorithms 
to describe continuous spatial patterns of divergence 
using gradients of admixture between clusters.
For example, in Fig \ref{bear_K3}b, 
the third cluster (in gold) exhibits a clear East-West gradient that 
overlays the discrete structure between the Southwest cluster and the Southeast.
The results from ADMIXTURE are are not identical to 
those obtained using the nonspatial \texttt{conStruct} model, 
but they do show the same tendency: 
e.g., at $K=3$ 
-- the preferred model from the cross-validation 
analysis shown in Fig \ref{bear_admix_CVerrors} -- 
ADMIXTURE splits the westernmost Alaskan samples 
out of the cluster with the eastern samples, 
and at $K=4$, it subdivides the eastern cluster 
into two geographically partitioned groups (Fig \ref{bear_admixture}).
Interestingly, for the nonspatial model implemented in ADMIXTURE, 
the preferred model has a smaller $K$ ($K=3$) than that of the 
spatial models with best cross-validation performance in \texttt{conStruct} ($K=5$ or 6). 
This discrepancy may be the result of differences in the cross-validation procedures 
implemented in \texttt{conStruct} and ADMIXTURE, 
or perhaps in the ways each accommodate missing data.

Comparing results of the two methods, it is clear that even at $K=3$, 
ADMIXTURE is invoking clusters to describe what seems to be, 
and is well described by \texttt{conStruct} as, 
a continuous spatial pattern of genetic variation. 
The third cluster in the ADMIXTURE analysis at $K=3$ 
(shown in gold in Fig \ref{bear_admixture}b), 
shows strong spatial autocorrelation in admixture proportions, 
as would be expected if it is describing continuous spatial differentiation.

@@

The proposed method combines the utility of model-based clustering algorithms 
with a biologically realistic model of isolation by distance.
We anticipate that \texttt{conStruct} will be useful for identifying populations 
and determining samples' ancestry in and across them, 
especially when the populations exhibit spatial patterns of relatedness.

\paragraph{Comparison to nonspatial model-based clustering}
Above, we showed that 
(a) the nonspatial \texttt{conStruct} model recapitulates results of 
other, commonly-used nonspatial clustering methods,
and 
(b) \texttt{conStruct} can concisely capture spatial structure, which
will be common within populations.
Given this, when should methods without spatial capability be used?
One advantage these have over \texttt{conStruct} is speed when the number of samples is large.
Although \texttt{conStruct}'s computation time is independent
of the number of loci included in the dataset 
(after the initial calculation of the allelic covariance), 
it currently scales poorly with number of samples.
The computationally limiting step is the inversion of the parametric covariance, 
which scales more than quadratically with the number of samples,
whereas computation time for, e.g., STRUCTURE, 
scales linearly with number of samples.
Our speed, on datasets with large sample sizes, 
could be improved by adopting rank-one updates 
to the inverse of the covariance matrix 
(e.g., \cite{woodbury1950,sherman_morrison1950})
when updating a sample's admixture proportions, 
which alters only a single row/column of the covariance matrix. 
We have not implemented this yet, 
as it would likely mean losing the ability to do efficient,  
Hamiltonian Monte Carlo sampling of our parameters.

For a relatively small number of samples, 
\texttt{conStruct} can be much faster than existing nonspatial Bayesian clustering methods.
On a desktop machine, using a single 4.2 GHz Intel Core i7 processor, 
an analysis of the black bear dataset (78 samples, 21,000 loci) 
running \texttt{conStruct}'s spatial model with 4 layers for 5,000 MCMC iterations 
(which was more than sufficient for convergence) 
took 2.8 hours.
On the same machine and dataset, 
a fastSTRUCTURE \cite{fastStructure} run with $K=4$ took 90.2 hours.
However, for a large number of samples, 
fastSTRUCTURE might be faster.
And, for almost any dataset of any size, 
the maximum likelihood algorithm implemented in ADMIXTURE 
is also quite a bit faster than \texttt{conStruct}; 
running ADMIXTURE on the bear dataset over all values of $K$ from 1 to 7, 
including 50-fold cross-validation for each value of $K$, took only 6.6 minutes.

@@

From a biological perspective, 
it is important to stress that patterns of relatedness between individuals and populations 
are shaped by complex spatial and hierarchical processes.
All individuals within a species (and indeed, all individuals across all species), 
are related to one another in some way, 
and summarizing that relatedness with a single value of $K$ may be reductive or misleading.
We therefore encourage users to perform analyses across different values of $K$ and 
observe which layers split out at what levels 
(this is conceptually similar to taking successively shallower
cross-sections of the population phylogeny), 
and also to take the results of the proposed cross-validation procedure with a large grain of salt.
Calculating layer contributions may also be a useful heuristic, 
as it can reveal layers with statistical support but small biological import.

Although we believe our model adds spatial realism to the groups used by clustering methods,
it is important to note that the layers detected by our method 
do not necessarily correspond  to distinct, ancestral populations; 
nor does a non-zero admixture proportion indicate that admixture 
(i.e. gene flow) must have occurred. 
Both groupings and admixture proportions
should be viewed as hypotheses that should be subject to further testing
(see \cite{Falush:16} for an in-depth discussion of these points).

\paragraph{Implications for management and conservation}
Because isolation by distance is common, 
a likely result of applying \texttt{conStruct} to existing data is that some
populations previously identified using nonspatial clustering methods 
may be collapsed into each other.  
This ``lumping'' might better reflect biological reality, 
but may also have implications for management decisions and conservation policy, 
both of which are often predicated on the identification of discrete ``management units'' (MUs) 
identified using genetic data \cite{Moritz1994,Waples_1998,Moritz_etal_2002}.

It is therefore important to stress that individuals sampled from the same \texttt{conStruct} layer 
may be quite genetically diverged from one another, 
perhaps especially at loci underlying adaptive traits, 
and that a \texttt{conStruct} layer may still contain multiple distinct MUs worthy of independent protections.  
Alternatively, the inclusion of multiple MUs into a single \texttt{conStruct} layer 
may occur if these populations are currently 
(or were recently) exchanging migrants, 
and thus might emphasize the importance of maintaining habitat corridors between demes, 
or of implementing an integrated conservation plan across multiple demes within a layer.

\paragraph{Allelic or genetic covariance?}
The choice of allelic covariance, rather than genetic covariance,
was motivated by two considerations.
First, it captures within-population variance in a manner more suited to the modeling approach here,
which led to better performance on test data.
Second, it is less affected by sample configuration --
the genetic covariance is calculated after subtracting the mean from the entire sample,
which is more strongly affected by densely sampled locations.
However, note that allelic covariance is more affected by singleton sites
than the standard genetic covariance,
so it may be advisable to filter these prior to analysis
if they are likely to contain a large percentage of errors \cite{linck_battey2017}.

\paragraph{Caveats and considerations}
There are a few important caveats to consider in the interpretation of \texttt{conStruct} results. 
First, we have modeled allelic covariance within a layer as a spatial process.
Although there is flexibility built into the model about the shape of that covariance, 
inference may be misleading if the sampling geography departs radically from the way 
the sampled organisms disperse (or have dispersed) on their landscape.
For example, if we were to run a \texttt{conStruct} analysis using geographic distances between 
sampled individuals of greenish warblers \cite{Irwin2001} or \textit{Ensatina} salamanders \cite{wake_schneider1998} 
--- two canonical examples of rings species --- 
we might get misleading results.
This is because distance between locations on either side of the species' distributions
(across the Tibetan plateau and the Central Valley, respectively) 
is not representative of the path traversed in the coalescent of a pair of alleles sampled at those locations.

@@

Thanks to averaging over choice of alleles,
the within-population allelic variance in sample $i$, $\widehat{\Omega}_{i,i}$,
is the variance of a series of Bernoulli($1/2$) draws across loci,
and therefore $\widehat{\Omega}_{i,i} = 1/4$ for every sample $i$.
Averaging over choice of reference allele therefore removes some information 
about factors acting within populations that might otherwise leave signatures in, 
e.g., the genetic covariance, 
such as population size, extent of inbreeding, and history of bottlenecks.  
However, as our model is focused on modeling 
\emph{co}variances between samples as the outcome of some spatial process, 
we count this a minor loss.

@@

A benefit of directly modeling the sample allelic covariance is that, 
after the initial calculation of the sample covariance matrix,
the computation time of the likelihood is not a function of the number of loci,
so inference can theoretically be done using whole genome data.

>>>>

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \cite{CavalliSforza1975, treemix},
dimensionality-reduction approaches, 
such as principal components analysis 
\cite{meirmans2009genodive,menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches 
(e.g., \cite{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, 
FINESTRUCTURE, fastStructure, huelsenbeck2007inference, 
Corander2003,TESS,geneland}).
Each of these methods perform best in particular situations,
but many can give misleading results when applied to data 
that show a continuous pattern of differentiation,
as that produced by geographic isolation by distance 
\cite{Wright1943, novembre_interpreting_2008, Frantz2009}.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
(We note that the problem of identifying population clusters is distinct from, 
though of course related to, 
the problem of detecting barriers to gene flow between populations, 
e.g.,  
\cite{BEDASSLE,EEMS,barton2008,Ringbauer2018}.)
Existing model-based clustering methods model each individual's genotypes 
as random draws from a set of underlying, unobserved population clusters, 
each with a characteristic set of allele frequencies, which are estimated. 
These underlying frequencies are identical 
for all individuals assigned to a cluster, 
\emph{regardless} of their spatial location. 
Spatial information has been incorporated into some of these methods, 
by, for example, placing spatial priors on cluster membership \cite{geneland,TESS}, 
but this does not address the underlying issue that 
these methods assume that allele frequencies 
are constant in a cluster across the species' range.

@@

\paragraph{Data}
The statistical framework of our approach is conceptually similar to that in \cite{Wasser2004}, \cite{BEDASSLE}, and \cite{spacemix},
although we use a somewhat different summary statistic than in this previous work.
The model works with allele frequencies at $L$ unlinked, bi-allelic single nucleotide polymorphisms (SNPs) genotyped across $N$ samples.
Each ``sample'' may be a single individual,  
a collection of individuals from a location, 
or frequencies estimated from pooled sequencing.
From these we compute the \emph{allelic covariance} 
between samples $i$ and $j$, denoted $\widehat{\Omega}_{i,j}$,
as the expected covariance of distinct individual alleles 
chosen from each of the two samples at a random locus.
More precisely, suppose that we pick a random bi-allelic locus uniformly from the genome,
pick a random ``reference'' allelic state from the two alleles seen at that locus,
and, in each sample, draw one random allele,
recording $X_i = 1$ if the allele drawn in sample $i$ matches the random reference,
and $X_i = 0$ otherwise.
Then, 
\begin{equation} \label{eqn:first_allelic_cov}
    \widehat{\Omega}_{i,j} = \cov[X_i, X_j] .
\end{equation}
Because we randomly choose the reference allele, 
each $X_i$ behaves marginally as a fair coin --
in particular, $\P\{X_i = 1\} = 1/2$, so $\widehat{\Omega}_{i,i} = 1/4$ for every $i$ --
all information enters through \emph{correlations}.

@@

Within each of these layers,
allele frequencies have positive covariance at geographically close locations,
but this covariance is allowed to decay as geographic distance increases.
This pattern of spatial decay reflects how migration between nearby spatial regions
homogenizes allele frequency changes that arise locally due to drift, 
but less effectively homogenizes geographically distant regions,
resulting in a continuous pattern of isolation by distance within each layer.
There is a fixed amount of covariance between layers, irrespective of spatial location.
Within each layer, allele frequencies are expected to change gradually with distance,
but observed frequencies can change abruptly at many loci 
if the proportions of ancestry individuals derive from each layer 
(the ``admixture'' proportions) 
do so as well.

@@

Cross-validation provides a valuable summary of how much explanatory power
is added by spatial structure within each layer, and each additional layer.
However, we remind users that ``statistical significance does not imply real-world significance'',
and so small but statistically significant differences between models
should not be relied on too strongly.

@@

To calculate layer contributions, we use the following alternative description of our covariance model:
the genomes of any pair of individuals \emph{agree} with some
background probability at a locus,
but this probability of agreement is increased on any segment of genome 
that both have inherited from the same layer 
(the amount it increases depends on how far apart they are
geographically and on the decay of isolation by distance).
We use this characterization to quantify the relative contributions of each layer,
by computing the average contribution to increased probability of agreement
as described in \secref{layer_contribution}. 
This layer contribution is similar to the ``ancestry contribution'' proposed by \cite{fastStructure}. 
However, each of our layers can induce a different amount of covariance between samples embedded in them, 
so we take that into account when calculating each layer's contribution to the whole.

@@

To test the method, we first generated data using the coalescent simulator \texttt{ms} \cite{Hudson2002}.
In each simulation,
we split a single ancestral population into $K$ subpopulations $\tau_{\text{s}}$ units of coalescent time in the past,
and at time $\tau_{\text{e}}$ in the past,
each of these discrete populations instantaneously colonized a
separate $6 \times 6$ square lattice of demes.
Migration on each lattice was to nearest neighbors (eight neighbors, including diagonals).
Finally, at time $\tau_{\text{a}}$ in the past, 
we collapsed those $K$ discrete layers into a single grid of demes,
choosing various amounts of admixture from these different layers
(see Fig \ref{sim_setup}). 
We collapsed the layers together using random, spatially autocorrelated admixture proportions 
(see \secref{sim_details}).
We simulated datasets using $K=1$, 2, and 3 layers;
in each simulation we sampled 10,000 unlinked loci from each of 20 haploid individuals from every deme.
We then ran both spatial and nonspatial \texttt{conStruct} analyses on each simulated dataset with $K$ between 1 and 7,
and compared predictive performance of the models using cross-validation with 10 replicates.
For comparison,
we also analyzed each simulated dataset using ADMIXTURE \cite{ADMIXTURE} with $K$ between 1 and 7, 
and compared models using ADMIXTURE's cross-validation procedure with 50 folds.

@@

The spatial model's better fit is reflected by increased predictive accuracy:
as shown in Fig \ref{sim_xvals},
across all models and choices of $K$, the spatial model is correctly preferred over the nonspatial model.
As desired, predictive accuracy of the spatial model increases until the true value of $K$,
and then plateaus or declines
(Figs \ref{sim_xvals} and \ref{simK1_xval}-\ref{simK3_xval}).
Predictive accuracy of the nonspatial model increases as subsequent clusters are added
up to $K=7$ (the largest number tested),
although gains are greatest as layers below the true number are added.
The same holds true for the ADMIXTURE cross-validation results,
in which models that have the largest number of clusters are preferred over all other models,
as shown in Fig \ref{sim_xvals} (vermilion diamonds),
and, in more detail, in Fig \ref{sim_admix_CVerrors}.

@@

\paragraph{Comparison to ADMIXTURE}
Both nonspatial \texttt{conStruct} and ADMIXTURE displayed 
the successive partitioning of space and the clines of admixture seen in the simulation results.
The details of each were somewhat different 
(Fig \ref{populus_nsp_pies} vs. Fig \ref{populus_admixture}), 
and also differed across the replicate analyses.
These differences between runs and methods may be due to 
noise in the different inference algorithms employed,
multi-modality in the likelihood surfaces, or to model details
(e.g., the priors used in nonspatial \texttt{conStruct}, 
or the fact that ADMIXTURE is modeling each allele's frequency in each cluster, 
rather than the covariance across all alleles).
However, overall, the behavior of both methods was quite similar: 
each recovered the \tri{}/\bals{} split with the first two clusters modeled, 
then, with higher values of $K$, 
used subsequent clusters to subdivide the \tri{} samples 
into geographically restricted foci of cluster membership.
Both nonspatial \texttt{conStruct} and ADMIXTURE 
strongly favored the most cluster-rich model
(Figs \ref{populus_xvals} and \ref{poplar_admix_CVerrors}).
In contrast, the spatial \texttt{conStruct} model clearly did not favor 
the model with the highest value of $K$, 
and appears to describe patterns of isolation by distance 
across the \tri{} range quite well.

@@

\paragraph{Results from \texttt{conStruct}}
The results partition the sampled bears into two main groups (shown in Fig \ref{bear_K3}a):
one (red) to the east of the Rocky Mountains, 
which also occurs in Alaska,
the other primarily west of the Rockies (blue).
The disjointed range of the blue layer likely reflects the fact
that Canada was not sampled, 
and so the red layer may extend
through the intervening (unsampled) northern Great Plains and Canadian Shield, 
with the blue layer presumably then stretching up into British Columbia.

The spatial models have strong statistical support 
up until around $K=5$ or 6 (Fig \ref{bear_xvals}),
but additional spatial layers beyond $K=2$ contribute little to total covariance (Fig \ref{bear_laycon}).
The locations of admixed individuals 
are consistent with a scenario of postglacial expansion from two refugia, 
one in the American Southwest and one in the American Southeast, 
meeting near the Northwest coast of North America and the Cascade Range.
However, lack of any samples from Canada and Mexico, 
and lack of denser sampling across northern North America, 
make more detailed interpretations untrustworthy.
The spatial covariance functions estimated in layers beyond the first two 
take very large values over small spatial lags, 
but decay sharply after that.
This feature, combined with the overall amounts 
and spatial patterns of ancestry in those layers, 
suggests that these layers are describing processes 
that shape genetic variation at local scales, such as inbreeding, 
which affects covariance between individuals within each location, 
but has limited impact on covariance between locations.

\paragraph{Comparison to ADMIXTURE}
Results from the nonspatial model and from the ADMIXTURE analyses 
clearly exhibit the tendency of nonspatial clustering algorithms 
to describe continuous spatial patterns of divergence 
using gradients of admixture between clusters.
For example, in Fig \ref{bear_K3}b, 
the third cluster (in gold) exhibits a clear East-West gradient that 
overlays the discrete structure between the Southwest cluster and the Southeast.
The results from ADMIXTURE are are not identical to 
those obtained using the nonspatial \texttt{conStruct} model, 
but they do show the same tendency: 
e.g., at $K=3$ 
--- the preferred model from the cross-validation 
analysis shown in Fig \ref{bear_admix_CVerrors} --- 
ADMIXTURE splits the westernmost Alaskan samples 
out of the cluster with the eastern samples, 
and at $K=4$, it subdivides the eastern cluster 
into two geographically partitioned groups (Fig \ref{bear_admixture}).
Interestingly, for the nonspatial model implemented in ADMIXTURE, 
the preferred model has a smaller $K$ ($K=3$) than that of the 
spatial models with best cross-validation performance in \texttt{conStruct} ($K=5$ or 6).
This discrepancy likely stems from the different features introduced in layers beyond $K=2$ in the two models:
conStruct uses small contributions of new layers to model very local drift,
while ADMIXTURE moves to geographically finer subdivisions.

Comparing results of the two methods, it is clear that even at $K=3$, 
ADMIXTURE invokes clusters to describe what seems to be, 
and is well described by \texttt{conStruct} as, 
a continuous spatial pattern of genetic variation. 
The third cluster in the ADMIXTURE analysis at $K=3$ 
(shown in gold in Fig \ref{bear_admixture}b), 
shows strong spatial autocorrelation in admixture proportions, 
as would be expected if it is describing continuous spatial differentiation.

@@

The proposed method combines the utility of model-based clustering algorithms 
with a biologically realistic model of isolation by distance.
We anticipate that \texttt{conStruct} will be useful for identifying populations 
and determining ancestral origins of sampled individuals,
especially when the populations exhibit geographic patterns of relatedness.

\paragraph{Comparison to nonspatial model-based clustering}
Above, we showed that 
(a) the nonspatial \texttt{conStruct} model recapitulates results of 
other, commonly-used nonspatial clustering methods,
and 
(b) \texttt{conStruct} can concisely capture spatial structure, which
is common within populations.
Given this, when should methods without spatial capability be used?
One advantage these have over \texttt{conStruct} is speed when the number of samples is large.
Although \texttt{conStruct}'s computation time is independent
of the number of loci included in the dataset 
(after the initial calculation of the allelic covariance), 
it currently scales poorly with number of samples.
The computationally limiting step is the inversion of the covariance matrix, 
which scales more than quadratically with the number of samples,
whereas computation time for, e.g., STRUCTURE, 
scales linearly with number of samples.
Our speed, on datasets with large sample sizes, 
could be improved by adopting rank-one updates 
to the inverse of the covariance matrix 
(e.g., \cite{woodbury1950,sherman_morrison1950})
when updating a sample's admixture proportions, 
which alters only a single row/column of the covariance matrix. 
We have not implemented this yet, 
as it would likely mean losing the ability to do efficient,  
Hamiltonian Monte Carlo sampling of our parameters.

For a relatively small number of samples, 
\texttt{conStruct} can be much faster than existing nonspatial Bayesian clustering methods.
On a desktop machine, using a single 4.2 GHz Intel Core i7 processor, 
an analysis of the black bear dataset (78 samples, 21,000 loci) 
running \texttt{conStruct}'s spatial model with 4 layers for 5,000 MCMC iterations 
(which was more than sufficient for convergence) 
took 2.8 hours.
On the same machine and dataset, 
a fastSTRUCTURE \cite{fastStructure} run with $K=4$ took 90.2 hours.
However, for a large number of samples, 
fastSTRUCTURE might be faster.
For almost any size dataset,
the maximum likelihood algorithm implemented in ADMIXTURE 
is quite a bit faster than \texttt{conStruct}:
running ADMIXTURE on the bear dataset over all values of $K$ from 1 to 7, 
including 50-fold cross-validation for each value of $K$, took only 6.6 minutes.

@@

From a biological perspective, 
it is important to stress that patterns of relatedness between individuals and populations 
are shaped by complex spatial and hierarchical processes.
All individuals within a species are related to one another in some way, 
and summarizing those patterns of relatedness with a single value of $K$ may be reductive or misleading.
We therefore encourage users to perform analyses across different values of $K$ and 
observe which layers split out at what levels 
(this is conceptually similar to taking successively shallower
cross-sections of the population phylogeny), 
and also to take the results of the proposed cross-validation procedure with a large grain of salt.
Calculating layer contributions may also be a useful heuristic, 
as it can reveal layers with statistical support but small biological import.

Although we believe our model adds spatial realism to the groups used by clustering methods,
it is important to note that the layers detected by our method 
do not necessarily correspond  to distinct, ancestral populations; 
nor does a non-zero admixture proportion indicate that admixture 
(i.e., gene flow) must have occurred. 
Both groupings and admixture proportions
should be viewed as hypotheses that should be subject to further testing
(see \cite{Falush:16} for an in-depth discussion of these points).

\paragraph{Implications for management and conservation}
Because isolation by distance is common, 
a likely result of applying \texttt{conStruct} to existing data is that
populations previously identified as distinct using nonspatial clustering methods 
may be grouped into the same layer.
This ``lumping'' might better reflect the demographic history of these populations,
but may not contradict the genetic distinctness implied by the nonspatial clustering.
This genetic distinctness rather than shared history may be more relevant for
management decisions and conservation policy,
which are often predicated on the identification of discrete ``management units'' (MUs) 
identified using genetic data \cite{Moritz1994,Waples_1998,Moritz_etal_2002}.

It is therefore important to stress that individuals sampled from the same \texttt{conStruct} layer 
may be quite genetically diverged from one another, 
perhaps especially at loci underlying adaptive traits, 
and that a \texttt{conStruct} layer may still contain multiple distinct MUs worthy of independent protections.  
For instance, although both the Alaskan and Eastern Black Bears draw most of their ancestry 
from the same conStruct layer, they are separated by a great distance, 
and may therefore differ substantially from each other
(although less than from the Western bears, as measured by average covariance).
Alternatively, the inclusion of multiple MUs into a single \texttt{conStruct} layer 
may occur if these populations are currently 
(or were recently) exchanging migrants, 
and thus might emphasize the importance of maintaining habitat corridors,
or of implementing an integrated conservation plan across a geographic region.

\paragraph{Allelic or genetic covariance?}
The choice of allelic covariance, rather than genetic covariance,
was motivated by the fact that
it is less affected by sample configuration --
the genetic covariance is calculated after subtracting the mean from the entire sample,
which is more strongly affected by densely sampled locations.
This led to better performance on test data.
However, note that allelic covariance is more affected by singleton sites
than the standard genetic covariance,
so it may be advisable to filter these prior to analysis
if they are likely to contain a large percentage of errors \cite{linck_battey2017}.

\paragraph{Caveats and considerations}
There are a few important caveats to consider in the interpretation of \texttt{conStruct} results. 
First, we have modeled allelic covariance within a layer as a spatial process.
Although there is flexibility built into the model about the shape of that covariance, 
inference may be misleading if the sampling geography departs radically from the way 
the sampled organisms disperse (or have dispersed) on their landscape.
For example, if we were to run a \texttt{conStruct} analysis using geographic distances between 
sampled individuals of greenish warblers \cite{Irwin2001} or \textit{Ensatina} salamanders \cite{wake_schneider1998} 
--- two canonical examples of ring species --- 
we might get misleading results.
This is because distance between locations on either side of the species' distributions
(across the Tibetan plateau and the Central Valley, respectively) 
is not representative of the path traversed in the coalescent of a pair of alleles sampled at those locations.

@@

Thanks to averaging over choice of alleles,
the within-population allelic variance in sample $i$, $\widehat{\Omega}_{i,i}$,
is the variance of a series of Bernoulli($1/2$) draws across loci,
and therefore $\widehat{\Omega}_{i,i} = 1/4$ for every sample $i$.
Averaging over choice of reference allele therefore removes some information 
about factors acting within populations that might otherwise leave signatures in
the genetic covariance, 
such as population size, extent of inbreeding, and history of bottlenecks.  
However, as our model is focused on modeling 
\emph{co}variances between samples as the outcome of some spatial process, 
we count this a minor loss.

@@

A benefit of directly modeling the sample allelic covariance is that, 
after the initial calculation of the sample covariance matrix,
the computation time of the likelihood is not a function of the number of loci,
so inference can be done using whole genome data.
